---
title: 'Spam detection example'
author:
- "Matt Taddy"
- "Itamar Caspi"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: haddock
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float: yes
abstract: |
  A replication of the spam detection example that appear in Matt Taddy's "Business Data Science"m Chapter 2 under "Logistic Regression."
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(eval = TRUE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

```

## Setup

Load required packages

```{r load_packages}

if (!require("pacman")) install.packages("pacman")

pacman::p_load(here,       # for referencing files and folders
               tidyverse,  # for data reading wrangling and visualization
               tidymodels) # for data modeling

```

Read the spam dataset

```{r read_data}

email <- read_csv(here("spam", "spam.csv"))

```

## Estimate the full model

we fit the full model (you don't need to worry about the warning. It says that some covariates are nearly perfect predictors.)

```{r estimate_model}

spammy <- glm(spam ~ ., data=email, family='binomial')

```

The guy is named George and he works in a cs dept
```{r george}

email %>% 
  group_by(spam, word_george, word_free) %>% 
  tally() %>% 
  pivot_wider(names_from = spam, values_from = n)

```

The model's coefficients
```{r glm_coeffs}

spammy %>% 
  tidy() %>% 
  filter(term %in% c("word_george", "word_free")) %>% 
  mutate(exp_estimate = exp(estimate),
         invexp_estimate = 1/exp(estimate)) %>% 
  select(term, exp_estimate, invexp_estimate)

```

Predict spam versus not-spam probabilities and odds-ratio for two arbitrary observations (the 1st and 4000th)

```{r}

# odds
spammy %>% 
  augment() %>% 
  select(.fitted) %>% 
  slice(1,4000)

# probability
spammy %>% 
  augment(type.predict = "response") %>% 
  select(.fitted) %>% 
  slice(1,4000)

```


Plot fitted values (probabilities) vs actual
```{r fitted_boxplot}

spammy %>% 
  augment(type.predict = "response") %>% 
  ggplot(aes(x = spam, y = .fitted, color = factor(spam))) +
  geom_boxplot() +
  labs(x = "",
       y = "Fitted probability of spam",
       color = "")

```


## Out-of-sample prediction

Split the data in two. One data set, called the training set, and the other, the test set. We do so, by using the `initial_split()`, `training()`, and `testing()` functions from the `rsampl` package:

```{r split}

set.seed(1234) # for the replicating the results

email_split <- email %>% 
  initial_split(prop = 0.7825) # the prop is set to make the size of the test set = 1000

email_split
```


```{r train_test}

email_training <- email_split %>% training()
email_testing  <- email_split %>% testing()

```


Estimate our model using the training set

```{r train}

spamtrain <- glm(spam ~ ., data=email_training, family='binomial')

```

Use the `spamtrain` model to make predictions on the test dataset

```{r predict}

spampredict <- spamtrain %>% 
  augment(newdata = email_testing,
          type.predict = "response")

```

Plot predicted values (probabilities) vs actual
```{r boxplot_predict}

spampredict %>% 
  ggplot(aes(x = spam, y = .fitted, color = factor(spam))) +
  geom_boxplot() +
  labs(x = "",
       y = "Predicted probability of spam",
       color = "")

```


## Deviance

In-sample proportion of variance explained ($R^2$)
```{r R2_insample}

glance(spammy) %>% 
  mutate(R2 = (1 - deviance / null.deviance)) %>% 
  select(null.deviance, deviance, R2)

```

Out-of-sample $R^2$ 
```{r log_loss}

# This function returns the log loss for a single prediction
log_loss <- function(actual, pred) {
  loss <- -2 * (actual * log(pred) + (1 - actual) * log(1 - pred))
  return(loss)
}

```

```{r R2_OOS}

spampredict %>% 
  mutate(.fitted_null = mean(spam),
         error = log_loss(spam, .fitted),
         error.null = log_loss(spam, .fitted_null)) %>%
  summarise(deviance = sum(error),
            deviance.null = sum(error.null)) %>% 
  mutate(R2 = 1 - deviance / deviance.null)
  
```

