---
title: 'Credit Score'
author:
- "Matt Taddy"
- "Itamar Caspi"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: haddock
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float: yes
abstract: |
  A replication of the German credit score example that appear in Matt Taddy's "Business Data Science"m Chapter 4 under "Probability, Cost, and Classification."
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(eval = TRUE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

```

## Load required packages

```{r load_packages}

if (!require("pacman")) install.packages("pacman")

pacman::p_load(here,        # for referencing files and folders
               tidyverse,   # for data reading wrangling and visualization
               tidymodels , # for data modeling
               ggmosaic,    # for mosaic plots
               gamlr)       # for running the gamma lasso algorithm
```


## Read and pre-process the data

Read the data
```{r read_data}

credit_raw <- read_csv(here("credit", "credit.csv"))

```

Create some "interesting" variables by re-leveling the credit history and checking account status
```{r preprocess}

credit_processed <- credit_raw %>%
  mutate(history = fct_collapse(history,
    good     = c("A30", "A31"),
    poor     = c("A32", "A33"),
    terrible = c("A34")
  )) %>%
  mutate(foreign = fct_recode(foreign,
    "foreign" = "A201",
    "german"  = "A202"
  )) %>%
  mutate(purpose = fct_collapse(purpose,
    newcar       = c("A40"),
    usedcar      = c("A41"),
    goods_repair = c("A42", "A43", "A44", "A45"),
    edu          = c("A46", "A48"),
    na           = c("A47"),
    biz          = c("A49", "A410")
  )) %>% 
  mutate(rent = factor(housing == "A151"))
```

Our analysis focuses on a subset of nine variables
```{r select}

credit_df <- credit_processed %>% 
  select(Default, duration, amount,
         installment, age, history,
         purpose, foreign, rent)
  
```

Plot a mosaic using the `ggmosaic` package
```{r mosaicplot}

credit_df %>% 
  mutate(Default = as_factor(Default)) %>% 
  ggplot() +
  geom_mosaic(aes(x = product(Default, history), fill = Default)) + 
  labs(x = "History",
       y = "Default",
       fill = "")

```
Surprise! the dangers of choice-based sampling!

## Build a design matrix


+ `recipe()` starts a new set of transformations to be applied, Its main argument is a formula that defines the outcome (`Default` in our case) and the predictors.  
+ `step_x()` defines data transformations.  
+ `prep()` prepares the transformations based on the data that is supplied. 
+ `juice()` to generate the outcome and predictors matrix to be used in the eatimation step
```{r recipe}

credit_recipe <- credit_df %>% 
  recipe(Default ~ .) %>% # defines Default as outcome and the rest as predictors
  step_dummy(all_nominal(), -Default,
             one_hot = TRUE) %>% # factors to dummies, keep all levels
  step_interact(~ all_predictors():all_predictors()) %>% # pairwise interactions
  step_zv(all_predictors()) %>% # remove redundent interactions
  prep()

credit_juiced <- juice(credit_recipe) 

head(credit_juiced)
```


Estimate the model using `cv.gamlr()`
```{r estimate_gamlr}

cred_X <- credit_juiced %>% select(-Default) # predictors matrix
cred_Y <- credit_juiced %>% select(Default)  # outcome vector

credscore <- cv.gamlr(x = cred_X, y = cred_Y,
                      family = "binomial",
                      verb = TRUE)
```


Plot the lasso's regularization path and the cross-validation error for each value of $\lambda$
```{r plot_gamlr}

plot(credscore$gamlr)
plot(credscore)

```

For the selected set of random folds, check how many variables are selected by each selection criterion
```{r info_criteria}

model_spec <- tribble(~criterion, ~var_selected,
  "min",  sum(coef(credscore, s = "min") != 0),
  "AICc", sum(coef(credscore$gamlr) != 0),
  "1se",  sum(coef(credscore) != 0),
  "AIC",  sum(coef(credscore$gamlr, s = which.min(AIC(credscore$gamlr))) != 0),
  "BIC",  sum(coef(credscore$gamlr, s = which.min(BIC(credscore$gamlr))) != 0))

model_spec

```

Augment the `credit_df` data with the model's prediction vector
```{r augment_gamlr}

pred <- predict(credscore$gamlr, newdata = cred_X, type="response") %>% 
  drop() # remove the sparse matrix formatting

credit_df_pred <- credit_df %>% 
  mutate(.pred_prob = pred) %>% 
  select(Default, .pred_prob)

head(credit_df_pred)
```

Show in sample fitted probability (via a boxplot)
```{r boxplot}

credit_df_pred %>% 
  ggplot(aes(x = as_factor(Default),
             y = .pred_prob,
             color = as_factor(Default))) +
  geom_boxplot() +
  labs(x = "Default",
       y = "Fitted probability of default",
       color = "Default")

```

## Misclassification rates

Classify predictions to "1" and "0" based on an arbitrary rule
```{r rule}

rule <- 0.20 # move this around to see how these change

credit_df_rule <- credit_df_pred %>% 
  mutate(.pred_class = case_when(.pred_prob >= rule ~ 1,
                                 .pred_prob <  rule ~ 0))

head(credit_df_rule)
```

What are the misclassification rates?
```{r accuracy}

options(yardstick.event_first = FALSE) # consider "1" as the "positive" result

sensitivity <- credit_df_rule %>%
  sens(factor(Default), factor(.pred_class)) %>% 
  pull(.estimate)

specificity <- credit_df_rule %>%
  spec(factor(Default), factor(.pred_class)) %>% 
  pull(.estimate)

accuracy_tbl <- tribble(~measure, ~value,
                        "false positive rate", 1 - specificity,
                        "false negative rate", 1 - sensitivity,
                        "sensitivity", sensitivity,
                        "specificity", specificity)

accuracy_tbl
```


## Out of sample Prediction

Split the sample in half
```{r split}

set.seed(1234) # for the replicating the results

credit_split <- credit_df %>% 
  initial_split(prop = 0.5) 

credit_split
```

Prepare the recipe to be used later to generate the training and test samples
```{r recipe_split}

credit_recipe <- training(credit_split) %>% 
  recipe(Default ~ .) %>% 
  step_dummy(all_nominal(), - Default, one_hot = TRUE) %>% 
  step_interact(~ all_predictors():all_predictors()) %>%
  step_zv(all_predictors()) %>% 
  prep()

```

Use the recipe to generate a training and test samples
```{r train_test}

credit_training <- credit_recipe %>% 
  juice()

credit_testing <- credit_recipe %>% 
  bake(new_data = testing(credit_split))

```

Train the model
```{r predict_oos}

cred_X_train <- credit_training %>% select(-Default)
cred_Y_train <- credit_training %>% select(Default)

credscore <- cv.gamlr(x = cred_X_train, y = cred_Y_train,
                      family = "binomial", verb = TRUE)

```

Evaluate the performance of of the model using the training set
```{r roc_in_sample}

pred_is <- predict(credscore$gamlr, newdata = cred_X_train, type="response") %>% 
  drop()

credit_training %>% 
  mutate(.pred_prob = pred_is) %>% 
  roc_curve(as_factor(Default), .pred_prob) %>% 
  autoplot() +
  labs(title = "In-sample ROC curve")

```

Evaluate the performance of the model using the test set
```{r roc_out_of_sample}

cred_X_test <- credit_testing  %>% select(-Default)

pred_oos <- predict(credscore$gamlr, newdata = cred_X_test, type="response") %>% 
  drop()

credit_testing %>%
  mutate(.pred_prob = pred_oos) %>% 
  roc_curve(as_factor(Default), .pred_prob) %>% 
  autoplot() +
  labs(title = "Out-of-sample ROC curve")


```

## References

Taddy, Matt. _Business Data Science: Combining Machine Learning and Economics to Optimize, Automate, and Accelerate Business Decisions_ . McGraw-Hill Education.